Modelling biological pathway dynamics\\with \tas

### 2013-06-28 Paper.tex
Updated red-yellow-green to red-yellow-cyan

This paper extends on the BIBE conference paper presented in 2012~\cite{animo-bibe}. In the present work, we describe in detail how 
\tas\ guards and invariants are calculated from kinetic reaction equations and how unwanted side effects of rounding 
to integer time units are dealt with. Furthermore, we show how the use of UPPAAL's statistical model checker~\cite{uppaal-smc} improves simulation 
performance. Finally, a newly implemented feature is presented that assists in understanding the effect of 
modifications to the model, by using a heatmap that shows the dynamic per-node differences between two versions of the model. 
To exemplify this new functionality, we provide an illustrative case study that shows the added value for performing \emph{in silico} experiments.

@@

\renewcommand{\tabcolsep}{2mm}
\begin{table}[htbp]
\caption{Lower and upper bounds for a reaction $A \rightarrow B$ with kinetic scenario 2, $k = 0.004$ (corresponding to the \emph{medium} preset)
and 10 seconds per UPPAAL time step. Granularity of $A$ is 5 (i.e. 5 activation steps from 0 to maximum activity), granularity of $B$ is 3. In both cases,
0 means completely inactive. With these settings, $\mbox{\it levelsScale} = 5.0$ and $\mbox{\it timeScale} = 0.1$. Uncertainty is set at $5 \%$.
{\color{BrickRed}Red} numbers are lower bounds, {\color{ForestGreen} green} numbers are upper bounds.}\label{tab:lower-upper-bounds}
\centering
    \begin{tabular}{|c||c|c|c|c|c|c|}
      \hline
      {\backslashbox[2em]{$B$\kern-2em}{\kern-1em$A$}} & \makebox[.5em]{0} & \makebox[.5em]{1} & \makebox[.5em]{2} & \makebox[.5em]{3} & \makebox[.5em]{4} & \makebox[.5em]{5} \\
      \hline\hline
      \begin{minipage}[c][1em][c]{1em}\centering 0\end{minipage} & \makebox[.5em]{$\infty$} & [{\scriptsize\color{BrickRed}40}, {\scriptsize\color{ForestGreen}44}] & [{\scriptsize\color{BrickRed}20}, {\scriptsize\color{ForestGreen}22}] & [{\scriptsize\color{BrickRed}13}, {\scriptsize\color{ForestGreen}15}] & [{\scriptsize\color{BrickRed}10}, {\scriptsize\color{ForestGreen}11}] & [{\scriptsize\color{BrickRed}8}, {\scriptsize\color{ForestGreen}9}] \\
      \hline
      \begin{minipage}[c][1em][c]{1em}\centering 1\end{minipage} & \makebox[.5em]{$\infty$} & [{\scriptsize\color{BrickRed}59}, {\scriptsize\color{ForestGreen}66}] & [{\scriptsize\color{BrickRed}30}, {\scriptsize\color{ForestGreen}33}] & [{\scriptsize\color{BrickRed}20}, {\scriptsize\color{ForestGreen}22}] & [{\scriptsize\color{BrickRed}15}, {\scriptsize\color{ForestGreen}16}] & [{\scriptsize\color{BrickRed}12}, {\scriptsize\color{ForestGreen}13}] \\
      \hline
      \begin{minipage}[c][1em][c]{1em}\centering 2\end{minipage} & \makebox[.5em]{$\infty$} & [{\scriptsize\color{BrickRed}119}, {\scriptsize\color{ForestGreen}131}] & [{\scriptsize\color{BrickRed}59}, {\scriptsize\color{ForestGreen}66}] & [{\scriptsize\color{BrickRed}40}, {\scriptsize\color{ForestGreen}44}] & [{\scriptsize\color{BrickRed}30}, {\scriptsize\color{ForestGreen}33}] & [{\scriptsize\color{BrickRed}24}, {\scriptsize\color{ForestGreen}26}] \\
      \hline
      \begin{minipage}[c][1em][c]{1em}\centering 3\end{minipage} & \makebox[.5em]{$\infty$} & $\infty$ & $\infty$ & $\infty$ & $\infty$ & $\infty$\\
      \hline
    \end{tabular}
\end{table}

@@

\begin{figure}[htb]
\centering
  \includegraphics[width=.48\textwidth]{mapk_model_diff4}
\caption{Difference between the simulations obtained using
50 ng/ml NGF and 100 ng/ml EGF as input at 22 minutes (vertical red bar in right panel). In this mode, node colours
are based on the difference of activity between the two configurations of the model:
a green-coloured node is more active due to NGF treatment, while red indicates less activity and white
means no difference. Note that NGF and EGF are both white, since at 22 minutes, 
NGF and EGF are inactivated in both conditions by the neutralizing antibody. A slider underneath the graph can be used to 
interactively view the differences at different time points. \label{fig:case-study-diff}}
\end{figure}

>>>>

This paper extends on the BIBE conference paper presented in 2012~\cite{animo-bibe}. In the present work, we describe in detail how 
\tas\ guards and invariants are calculated from kinetic reaction equations and how unwanted side effects of rounding 
to integer time units are dealt with. Furthermore, we show how the use of UPPAAL's statistical model checker~\cite{uppaal-smc} improves simulation 
performance. Finally, a newly implemented feature is presented that assists in understanding the effect of 
modifications to the model, by using a heat-map that shows the dynamic per-node differences between two versions of the model. 
To exemplify this new functionality, we provide an illustrative case study that shows the added value for performing \emph{in silico} experiments.

@@

\definecolor{ourBlue}{rgb}{0, 0.8, 0.8}
\renewcommand{\tabcolsep}{2mm}
\begin{table}[htbp]
\caption{Lower and upper bounds for a reaction $A \rightarrow B$ with kinetic scenario 2, $k = 0.004$ (corresponding to the \emph{medium} preset)
and 10 seconds per UPPAAL time step. Granularity of $A$ is 5 (i.e. 5 activation steps from 0 to maximum activity), granularity of $B$ is 3. In both cases,
0 means completely inactive. With these settings, $\mbox{\it levelsScale} = 5.0$ and $\mbox{\it timeScale} = 0.1$. Uncertainty is set at $5 \%$.
{\color{BrickRed}Red} numbers are lower bounds, {\color{ourBlue} cyan} numbers are upper bounds.}\label{tab:lower-upper-bounds}
\centering
    \begin{tabular}{|c||c|c|c|c|c|c|}
      \hline
      {\backslashbox[2em]{$B$\kern-2em}{\kern-1em$A$}} & \makebox[.5em]{0} & \makebox[.5em]{1} & \makebox[.5em]{2} & \makebox[.5em]{3} & \makebox[.5em]{4} & \makebox[.5em]{5} \\
      \hline\hline
      \begin{minipage}[c][1em][c]{1em}\centering 0\end{minipage} & \makebox[.5em]{$\infty$} & [{\scriptsize\color{BrickRed}40}, {\scriptsize\color{ourBlue}44}] & [{\scriptsize\color{BrickRed}20}, {\scriptsize\color{ourBlue}22}] & [{\scriptsize\color{BrickRed}13}, {\scriptsize\color{ourBlue}15}] & [{\scriptsize\color{BrickRed}10}, {\scriptsize\color{ourBlue}11}] & [{\scriptsize\color{BrickRed}8}, {\scriptsize\color{ourBlue}9}] \\
      \hline
      \begin{minipage}[c][1em][c]{1em}\centering 1\end{minipage} & \makebox[.5em]{$\infty$} & [{\scriptsize\color{BrickRed}59}, {\scriptsize\color{ourBlue}66}] & [{\scriptsize\color{BrickRed}30}, {\scriptsize\color{ourBlue}33}] & [{\scriptsize\color{BrickRed}20}, {\scriptsize\color{ourBlue}22}] & [{\scriptsize\color{BrickRed}15}, {\scriptsize\color{ourBlue}16}] & [{\scriptsize\color{BrickRed}12}, {\scriptsize\color{ourBlue}13}] \\
      \hline
      \begin{minipage}[c][1em][c]{1em}\centering 2\end{minipage} & \makebox[.5em]{$\infty$} & [{\scriptsize\color{BrickRed}119}, {\scriptsize\color{ourBlue}131}] & [{\scriptsize\color{BrickRed}59}, {\scriptsize\color{ourBlue}66}] & [{\scriptsize\color{BrickRed}40}, {\scriptsize\color{ourBlue}44}] & [{\scriptsize\color{BrickRed}30}, {\scriptsize\color{ourBlue}33}] & [{\scriptsize\color{BrickRed}24}, {\scriptsize\color{ourBlue}26}] \\
      \hline
      \begin{minipage}[c][1em][c]{1em}\centering 3\end{minipage} & \makebox[.5em]{$\infty$} & $\infty$ & $\infty$ & $\infty$ & $\infty$ & $\infty$\\
      \hline
    \end{tabular}
\end{table}

@@

\begin{figure}[htb]
\centering
  \includegraphics[width=.48\textwidth]{mapk_model_diff4}
\caption{Difference between the simulations obtained using
50 ng/ml NGF and 100 ng/ml EGF as input at 22 minutes (vertical red bar in right panel). In this mode, node colours
are based on the difference of activity between the two configurations of the model:
a cyan-coloured node is more active due to NGF treatment, while red indicates less activity and white
means no difference. Note that NGF and EGF are both white, since at 22 minutes, 
NGF and EGF are inactivated in both conditions by the neutralizing antibody. A slider underneath the graph can be used to 
interactively view the differences at different time points. \label{fig:case-study-diff}}
\end{figure}
